Verifying a Binary Search Tree with Fine-Grained Locking

### 2021-04-13 concurrency/Verification of Concurrent Binary Search Tree.tex
citation updated

\catcode`\@=11
\section{Introduction}
Modern CPU design trends have shifted towards multicore processing over the last two decades. This is because the computing power of computers has reached its limit in single-threaded execution, while the quantity of transistors per single-core CPU are increasing as described by Moore's law \cite{591665}. CPUs have multiple cores now because that is the only way to get more transistors per chip, and achieve a desirable throughput. However, doubling the quantity of cores on one chip does not immediately imply an increase in the performance by a factor of two. Programs that we aim to run in multiple cores have to be carefully designed to efficiently divide the workload into threads, and facilitate non-conflicting access to the shared data by multiple threads in parallel. These challenges is resolved by designing the concurrent data structures, algorithms that provide thread-safe access to the shared data by storing and organizing data in a standard way. Concurrent data structures are now widely employed in real-world applications like database algorithms, file systems, networking, e-commerce, etc. Designing these data structures has been considered as one of the difficult tasks in the software development fields. Despite the careful design, implementation, and thorough unit testing of these algorithms, it is very likely to expect subtle bugs and unexpected behaviors of the program. This is because threads interact with each other in complex and subtle ways such that it is almost impossible to cover every corner case while testing. Therefore, it is important to use a precise way of proving the correctness of concurrent data structures which provides a guarantee that the given algorithm correctly implements the specified concurrent data structures.

@@

In this thesis, we describe the formal verification of the correctness of concurrent binary search tree with fine-grained locking. We implement the BST algorithms in C and use the Verified Software Toolchain(VST) \cite{10.5555/2670099} to verify with respect to given specification. The concurrent separation logic (CSL) is employed to reason about behavior of the program implemented in the C programming language. Separation logic (SL) \cite{1029817} is an extension of Hoare logic \cite{10.1145/363235.363259}, way of reasoning about programs using a set of logical rules. SL helps to locally reason about programs that play with pointer data structures, where specifications and proofs of a program component only reflect the chunk of memory that the component uses, not the entire global state of the system. Concurrent separation logic (CSL) is a variant of the separation logic for reasoning about the concurrent programs. The version of CSL formalized in VST is rooted on the work of Hobor et al. \cite{10.1007/978-3-540-78739-6_27}, but extended with higher-order ghost state in the style of Iris \cite{10.1145/2951913.2951943}.
\section{Background}
\subsection{VST and Iris}
We use the Verified Software Toolchain \cite{10.5555/2670099}  to prove the correctness of the concurrent binary search tree. VST is a Coq-based proof system that helps to prove separation logic properties of C programs with the assurance that the properties will be preserved in compiled code. First, we write programs in C programming and using CompCert’s \emph{clightgen}, the ASTs for C programs can be generated in Coq. We then state and prove specifications for them using Verifiable C. Verifiable C is a higher-order separation logic for reasoning about the functional correctness of C programs and carries a soundness proof that links high-level specifications to assembly code generated by the CompCert Verified compiler. Verifiable C supports reasoning about two separate kinds of concurrency:  Pthreads-style concurrency with locks (used by implementing blocking data structures that can be fine-grained or coarse-grained ), and concurrency using C11 atomic operations (i.e. lock-free data structures).

\subsection{Ghost State and Global Invariants}
We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program. But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness. For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads. But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.   
\begin{figure}[htb]
\centering
$\texttt{x = 0;}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\hspace{30px}\pi = \pi_1\ .\ \pi_2$\\
$\begin{array}{l || l}
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{acquire(l);} & \texttt{acquire(l);}\\
\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{x++;} & \texttt{x++;}\\
\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{release(l);} & \texttt{release(l);}\\
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\end{array}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
\caption{The increment example annotated with separation logic assertion}
\label{figure1}
\end{figure}
We must preserve the connection between the value of shared resources and the work performed by each thread. A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread. They do not appear in the original program. Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}. In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together. VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} (\textsf{gname} in Coq) associated with the value $a$, which may be of any type. We will see different kind of ghost states used to verify binary search tree in Chapter \ref{chapter4}.

The main idea behind the logic from Iris \cite{10.1145/3022670.2951943}, a mechnized higher-order concurrent separation logic framework, is the construction of \emph{global invariant} as \emph{ghost state}. \emph{Global Invariant} is an invariant on the global (ghost and physical) state of program. We can open any invariant but must close it again before taking any steps of execution unless those execution are atomic. Thread can use the contents of \emph{global invariant} during atomic operation if it can guarantee that no one will ever see an intermediate state in which invariant does not hold. The rules from Jung et al.\cite{10.1145/3022670.2951943} for creating and opening invariants are:

@@

\section{Correctness Proofs}
In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct. To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant. We can accomplish this with ghost variables, a simple form of auxiliary state. Aside from the \emph{permission} on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the \emph{ghost state}. In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together. We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.

@@

The $\mathsf{sync\_inv}$ predicate associates the contribution part of ghost state $a$ for a node with ghost name $g$, along with the other concrete information about a node parameterized on $a$. In the atomic specification, we protect the $\mathsf{sync\_inv}$ using lock inside lock invariant. So, once we acquire the lock, we get access to $\mathsf{sync\_inv}$. The $\mathsf{sync\_commit}$ and $\mathsf{sync\_rollback}$, specialized with ghost state with \emph{reference pattern} ($\mathsf{my\_half}$ and $\mathsf{public\_half}$), are the versions of $\mathsf{atomic\_commit}$ and $\mathsf{atomic\_rollback}$ described in Chapter \ref{chapter2}. In a commit, we must provide $\mathsf{sync\_inv}$ with resources $R$ that, in combination with the public precondition, allow us to prove the public postcondition; we then gain access to an assertion Q required by the postcondition of the function, and leftover resources $R'$. The state of the node $g$ in public precondition (described by an assertion $\mathsf{public\_half}$) must be joinable with the state described by an assertion $\mathsf{sync\_inv}$. In a rollback, we provide $\mathsf{sync\_inv}$ with resources $R$ that, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as $\mathsf{sync\_inv}$ with the leftover resources $R$. We must use these rules before releasing the node's lock (we lose the $\mathsf{sync\_inv}$ after releasing the lock).

@@

The \emph{lock}, used in the implementation of functions, provides \texttt{acquire} and \texttt{release} functions, which allow threads to transfer ownership of resource invariants. Their pre- and postconditions are as follows:
$$\{!!\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{acquire}(\ell)\ \{R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
$$\{!!(\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{exclusive}\ R) * R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{release}(\ell)\ \{\mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$

@@

\section{Related Work}
Our work is based on the concurrent separation logic of VST in the style of Iris \cite{10.1145/3022670.2951943} and the logical atomicity from TaDA logic  \cite{tada}. The notion of the \emph{lower} and \emph{upper} bound is taken from the flow interface paper \cite{krishna2017flow}. Our main technical contributions are the verification of two different implementation of binary search tree (using fine-grained locking and lock-free technique) with respect to the same abstract specification, and first c-level mechanized verification of concurrent search-based data structure (i.e. BST). Most of the other concurrent separation logic (CSL)-based verification works are on toy languages, while VST lets us use the same logic on real C programs.

@@

%% Bibliography
\bibliography{sources}

\end{document}

>>>>

\catcode`\@=11
\section{Introduction}
Modern CPU design trends have shifted towards multicore processing over the last two decades. This is because the computing power of computers has reached its limit in single-threaded execution, while the quantity of transistors per single-core CPU are increasing as described by Moore's law \cite{moore}. CPUs have multiple cores now because that is the only way to get more transistors per chip, and achieve a desirable throughput. However, doubling the quantity of cores on one chip does not immediately imply an increase in the performance by a factor of two. Programs that we aim to run in multiple cores have to be carefully designed to efficiently divide the workload into threads, and facilitate non-conflicting access to the shared data by multiple threads in parallel. These challenges is resolved by designing the concurrent data structures, algorithms that provide thread-safe access to the shared data by storing and organizing data in a standard way. Concurrent data structures are now widely employed in real-world applications like database algorithms, file systems, networking, e-commerce, etc. Designing these data structures has been considered as one of the difficult tasks in the software development fields. Despite the careful design, implementation, and thorough unit testing of these algorithms, it is very likely to expect subtle bugs and unexpected behaviors of the program. This is because threads interact with each other in complex and subtle ways such that it is almost impossible to cover every corner case while testing. Therefore, it is important to use a precise way of proving the correctness of concurrent data structures which provides a guarantee that the given algorithm correctly implements the specified concurrent data structures.

@@

In this thesis, we describe the formal verification of the correctness of concurrent binary search tree with fine-grained locking. We implement the BST algorithms in C and use the Verified Software Toolchain(VST) \cite{plfcc} to verify with respect to given specification. The concurrent separation logic (CSL) is employed to reason about behavior of the program implemented in the C programming language. Separation logic (SL) \cite{seplogic} is an extension of Hoare logic \cite{hoare}, way of reasoning about programs using a set of logical rules. SL helps to locally reason about programs that play with pointer data structures, where specifications and proofs of a program component only reflect the chunk of memory that the component uses, not the entire global state of the system. Concurrent separation logic (CSL) is a variant of the separation logic for reasoning about the concurrent programs. The version of CSL formalized in VST is rooted on the work of Hobor et al. \cite{oraclesematic}, but extended with higher-order ghost state in the style of Iris \cite{higherorderghoststate}.
\section{Background}
\label{background}
\subsection{VST and Iris}
We use the Verified Software Toolchain \cite{plfcc}  to prove the correctness of the concurrent binary search tree. VST is a Coq-based proof system that helps to prove separation logic properties of C programs with the assurance that the properties will be preserved in compiled code. First, we write programs in C programming and using CompCert’s \emph{clightgen}, the ASTs for C programs can be generated in Coq. We then state and prove specifications for them using Verifiable C. Verifiable C is a higher-order separation logic for reasoning about the functional correctness of C programs and carries a soundness proof that links high-level specifications to assembly code generated by the CompCert Verified compiler. Verifiable C supports reasoning about two separate kinds of concurrency:  Pthreads-style concurrency with locks (used by implementing blocking data structures that can be fine-grained or coarse-grained ), and concurrency using C11 atomic operations (i.e. lock-free data structures).

\subsection{Ghost State and Global Invariants}
We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program. But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness. For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads. But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.   
\begin{figure}[htb]
\centering
$\texttt{x = 0;}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\hspace{30px}\pi = \pi_1\ .\ \pi_2$\\
$\begin{array}{l || l}
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{acquire(l);} & \texttt{acquire(l);}\\
\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{x++;} & \texttt{x++;}\\
\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{release(l);} & \texttt{release(l);}\\
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\end{array}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
\caption{The increment example annotated with separation logic assertion}
\label{figure1}
\end{figure}
We must preserve the connection between the value of shared resources and the work performed by each thread. A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread. They do not appear in the original program. Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}. In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together. VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} (\textsf{gname} in Coq) associated with the value $a$, which may be of any type. We will see different kind of ghost states used to verify binary search tree in Section \ref{correctness}.

The main idea behind the logic from Iris \cite{higherorderghoststate}, a mechnized higher-order concurrent separation logic framework, is the construction of \emph{global invariant} as \emph{ghost state}. \emph{Global Invariant} is an invariant on the global (ghost and physical) state of program. We can open any invariant but must close it again before taking any steps of execution unless those execution are atomic. Thread can use the contents of \emph{global invariant} during atomic operation if it can guarantee that no one will ever see an intermediate state in which invariant does not hold. The rules from Jung et al.\cite{higherorderghoststate} for creating and opening invariants are:

@@

 The \emph{lock}, used in the implementation of BST operations, provides \texttt{acquire} and \texttt{release} functions, which allow threads to transfer ownership of resource invariants. Their pre- and postconditions are as follows:
$$\{!!\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{acquire}(\ell)\ \{R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$
$$\{!!(\mathsf{readable\_share}\ \mathit{sh} \land \mathsf{exclusive}\ R) * R * \mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}\ \texttt{release}(\ell)\ \{\mathsf{lock\_inv}\ \mathit{sh}\ \ell\ R\}$$

\section{Correctness Proofs}
\label{correctness}
In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct. To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant. We can accomplish this with ghost variables, a simple form of auxiliary state. Aside from the \emph{permission} on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the \emph{ghost state}. In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together. We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.

@@

The $\mathsf{sync\_inv}$ predicate associates the contribution part of ghost state $a$ for a node with ghost name $g$, along with the other concrete information about a node parameterized on $a$. In the atomic specification, we protect the $\mathsf{sync\_inv}$ using lock inside lock invariant. So, once we acquire the lock, we get access to $\mathsf{sync\_inv}$. The $\mathsf{sync\_commit}$ and $\mathsf{sync\_rollback}$, specialized with ghost state with \emph{reference pattern} ($\mathsf{my\_half}$ and $\mathsf{public\_half}$), are the versions of $\mathsf{atomic\_commit}$ and $\mathsf{atomic\_rollback}$ described in Section \ref{background}. In a commit, we must provide $\mathsf{sync\_inv}$ with resources $R$ that, in combination with the public precondition, allow us to prove the public postcondition; we then gain access to an assertion Q required by the postcondition of the function, and leftover resources $R'$. The state of the node $g$ in public precondition (described by an assertion $\mathsf{public\_half}$) must be joinable with the state described by an assertion $\mathsf{sync\_inv}$. In a rollback, we provide $\mathsf{sync\_inv}$ with resources $R$ that, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as $\mathsf{sync\_inv}$ with the leftover resources $R$. We must use these rules before releasing the node's lock (we lose the $\mathsf{sync\_inv}$ after releasing the lock).

@@

@@

\section{Related Work}
Our work is based on the concurrent separation logic of VST in the style of Iris \cite{higherorderghoststate} and the logical atomicity from TaDA logic  \cite{tada}. The notion of the \emph{lower} and \emph{upper} bound is taken from the flow interface paper \cite{krishna2017flow}. Our main technical contributions are the verification of two different implementation of binary search tree (using fine-grained locking and lock-free technique) with respect to the same abstract specification, and first c-level mechanized verification of concurrent search-based data structure (i.e. BST). Most of the other concurrent separation logic (CSL)-based verification works are on toy languages, while VST lets us use the same logic on real C programs.

@@

%% Bibliography
\bibliography{sources}
\end{document}
